A Constraint Solving Approach to Parikh Images of Regular Languages

### 2024-03-07 calculus.tex
small stuff
Make automatas bigger, fix incorrect claim

We here only define forward propagation. Backward propagation 
can be straightforwardly obtained by computing reachability from
an accepting state on the same principle as above, and indeed our
implementation of \Calculus{} does this.

@@

In addition to \cref{tbl:rules:single}, we assume the existence of a rule
\PresburgerClose{}, corresponding to a sound and complete solver for Presburger
formulae, and for constraints over~$\Monoid$.

>>>>

@@

In addition to \cref{tbl:rules:single}, we assume the existence of a rule
\PresburgerClose{}, corresponding to a sound and complete solver for Presburger
arithmetic formulas, and for constraints over~$\Monoid$.

### 2024-03-07 data.tex
Add data statement
Make automatas bigger, fix incorrect claim

A reproduction package featuring all relevant logs is available~\cref{artefact}.
Additionally, \Ostrich{} and \Catra{} are both available as living software
on GitHub at \url{https://github.com/amandasystems/catra} and \url{https://github.com/uuverifiers/ostrich}
respectively. A version of \Ostrich{} modified to use CA-Str is available in the \texttt{cea-catra}~branch
and comes bundled with the reproduction package.

All source code to generate the artefact and the paper itself is \emph{also} available on 
GitHub at~\url{https://github.com/amandasystems/oopsla-artefact}.

\Ostrich{} and \Catra{} are both open source under a 3-clause BSD~license.

>>>>

A reproduction package featuring all relevant logs and all benchmarked instances is available~\cref{artefact}.
\Ostrich{} and \Catra{} are both available as living software
under a 3-clause BSD~license on GitHub at \url{https://github.com/amandasystems/catra} and \url{https://github.com/uuverifiers/ostrich}
respectively. A version of \Ostrich{} modified to use CA-Str is available in the \texttt{cea-catra}~branch. All source code to generate the paper itself
is available at~\url{https://github.com/amandasystems/oopsla-artefact}.

### 2024-03-07 example.tex
Make automatas bigger, fix incorrect claim

\newcommand{\autscale}[0]{0.48}

\begin{figure}[ht]
    \centering 
  \begin{subfigure}[b]{\autscale\textwidth}
    \centering
    \includegraphics[width=\textwidth]{a}
    \caption{The automaton $A'$, whose Parikh registers count the length of the
    string accepted by the automaton (register $l_A$), the start offset of the
    substring ($i$) and length ($n$) of the substring matching
    $\Regex{c\RegexOr{}dd}$. Note the symbolic transitions in the starting and
    accepting states matching any symbol in the alphabet!}\label{fig:aut_a}
      \Description[A 3-state Parikh automaton.]{%
      A 3-state Parikh automaton with two registers counting the length
      of a substring matching c or dd.%
      }
  \end{subfigure}%
  \hfill%
  \begin{subfigure}[b]{\autscale\textwidth}
    \centering
    \includegraphics[width=\textwidth]{b}
    \caption{The automaton $B$, where the Parikh registers count the length of
    the string in register $l_B$.}\label{fig:aut_b}
      \Description[A 4-state Parikh automaton.]{%
      A 4-state Parikh automaton with initial state IB going to the B state,
      which counts a number of letters c, followed by one final c before reaching 
      the accepting state, FB. The second path from the initial state consumes an a,
      goes into state A, consumes some number of c:s, followed by a final a to get to
      the accepting state.}
  \end{subfigure}
  \caption{The collection of automata we use as running
  examples, both derived from \cref{ex:string-constraints}.}\label{fig:examples}
\end{figure}

@@

We will now proceed to show how our calculus~\Calculus{} can lazily
prove the unsatisfiability of the running example. The calculus
interleaves several reasoning principles, which we will later
define precisely as calculus rules:
\begin{inparaenum}[(i)]
\item Similarly, as in \cite{generate-parikh-image}, we first describe
  each automaton as a flow network, counting how often each transition
  is taken in a run of the automaton. The flow constraints are an
  over-approximation of the possible accepting runs of an automaton.
\item We use a linear integer solver to simplify the flow constraints
  and prune away paths through the automata that are not feasible.
\item We use a tailor-made propagation algorithm to identify disconnected
  transitions of the automata that can never be taken and lazily add
  the corresponding constraints.
\item When propagation cannot infer further constraints, we use case
  splitting to subdivide the problem into smaller parts.
\item Once the constituent automata of a product have become
  sufficiently small, we compute the precise product automaton.
\end{inparaenum}

@@

\begin{figure}[tb]
  \centering 
  \includegraphics[width=0.6\textwidth]{a_2}
  \caption{ $\mathcal{A}'$ after even more simplification using linear algebra,
   with most transitions expressed
  in terms of $n$. Intuitively, this version captures the fact that $n$ is given
  by distributing the incoming flow of $1$ across the two outgoing transitions
  from $I_{A'}$, the initial state.}
  \Description[A 3-state automaton, with equations on its labels.]{%
  The initial state accepts any character and is labelled with the variable i. From
  that state there is a transition to the D state reading d with $n-1$ and one reading c
  going directly to the final state. The second transition is labelled with the equation
  $2-n$. The final state has one transition that consumes any character with the equation 
  $l_A-i-n$.%
  }
  \label{fig:a_2}
\end{figure}

\begin{figure}[ht]
  \centering 
\begin{subfigure}{0.42\textwidth}
  \centering
  \includegraphics[width=\textwidth]{a_annotated}
  \vspace*{5ex}
  \caption{ $\mathcal{A}'$ with its associated transition variables in symbolic
  form.}\label{fig:aut_a_annotated}
    \Description[A 3-state automaton with equations on its edges]{
      A 3-state automaton with a self-loop on any character on the
      initial state labelled 1, a transition on c to the accepting state
      and one on d to a state named D, labelled 0 and 1 respectively,
      then one transition from D to FA, the accepting state on d labelled 1.
      The accepting state has a self-loo labelled $l_A -3$.}
  \vspace*{1.9ex}
\end{subfigure}\hfill%
\begin{subfigure}{0.54\textwidth}
  \centering
  \includegraphics[width=\textwidth]{b_annotated}
  \caption{$\mathcal{B}$ with its associated transition variables in symbolic forms.
  The large number of implicitly existentially quantified
  variables~$x_i$ on
  transitions suggest that this automaton has a more complex structure
  with relation to its (free) target variable representing the string length
  than $\mathcal{A}'$. %The variables used here are fresh, and do not correspond to the
  % ones previously used.
}\label{fig:aut_b_annotated}
    \Description[A 4-state automaton with its transitions annotated
    with equations]{A 4-state automaton with transitions from the initial state to B on b,
    one to A on a, labelled $x_{12}$ and $x_{11}$ respectively. State B has one self-loop
    on c labelled $x_9$, one transition back to the initial state labelled 
    b / $x_{11} + x_{12} - 1$. Finally, B has a transition to the accepting state on c,
    labelled $1 0 x_{11}$. A has one self-loop on c labelled $l_A - x_9 - 2x_{11} - 2x_{12}$,
    and one transition on a to the accepting state labelled $x_{11}$.}
\end{subfigure}
\caption{The automata after arithmetic simplification.}\label{fig:propagated}
\end{figure}

>>>>

\newcommand{\autscale}[0]{0.60}

\begin{figure}[ht]
    \centering 
  \begin{subfigure}[b]{0.9\textwidth}
    \centering
    \includegraphics[width=\textwidth]{a}
    \caption{The automaton $A'$, whose Parikh registers count the length of the
    string accepted by the automaton (register $l_A$), the start offset of the
    substring ($i$) and length ($n$) of the substring matching
    $\Regex{c\RegexOr{}dd}$. Note the symbolic transitions in the starting and
    accepting states matching any symbol in the alphabet!}\label{fig:aut_a}
      \Description[A 3-state Parikh automaton.]{%
      A 3-state Parikh automaton with two registers counting the length
      of a substring matching c or dd.%
      }
  \end{subfigure}%
  \hfill%
  \begin{subfigure}[b]{\autscale\textwidth}
    \centering
    \includegraphics[width=\textwidth]{b}
    \caption{The automaton $B$, where the Parikh registers count the length of
    the string in register $l_B$.}\label{fig:aut_b}
      \Description[A 4-state Parikh automaton.]{%
      A 4-state Parikh automaton with initial state IB going to the B state,
      which counts a number of letters c, followed by one final c before reaching 
      the accepting state, FB.\@The second path from the initial state consumes an a,
      goes into state A, consumes some number of c:s, followed by a final a to get to
      the accepting state.}
  \end{subfigure}
  \caption{The collection of automata we use as running
  examples, both derived from \cref{ex:string-constraints}.}\label{fig:examples}
\end{figure}

@@

We will now proceed to show how our calculus~\Calculus{} can lazily
prove the unsatisfiability of the running example. The calculus
interleaves several reasoning principles, which we will later
define precisely as calculus rules:
\begin{inparaenum}[(i)]
\item Similarly, as in~\cite{generate-parikh-image}, we first describe
  each automaton as a flow network, counting how often each transition
  is taken in a run of the automaton. The flow constraints are an
  over-approximation of the possible accepting runs of an automaton.
\item We use a linear integer solver to simplify the flow constraints
  and prune away paths through the automata that are not feasible.
\item We use a tailor-made propagation algorithm to identify disconnected
  transitions of the automata that can never be taken and lazily add
  the corresponding constraints.
\item When propagation cannot infer further constraints, we use case
  splitting to subdivide the problem into smaller parts.
\item Once the constituent automata of a product have become
  sufficiently small, we compute the precise product automaton.
\end{inparaenum}

@@

\begin{figure}[tb]
  \centering 
  \includegraphics[width=0.9\textwidth]{a_2}
  \caption{ $\mathcal{A}'$ after even more simplification using linear algebra,
   with most transitions expressed
  in terms of $n$. Intuitively, this version captures the fact that $n$ is given
  by distributing the incoming flow of $1$ across the two outgoing transitions
  from $I_{A'}$, the initial state.}
  \Description[A 3-state automaton, with equations on its labels.]{%
  The initial state accepts any character and is labelled with the variable i. From
  that state there is a transition to the D state reading d with $n-1$ and one reading c
  going directly to the final state. The second transition is labelled with the equation
  $2-n$. The final state has one transition that consumes any character with the equation 
  $l_A-i-n$.%
  }
  \label{fig:a_2}
\end{figure}

\begin{figure}[ht]
  \centering 
\begin{subfigure}{0.42\textwidth}
  \centering
  \includegraphics[width=\textwidth]{a_annotated}
  \vspace*{5ex}
  \caption{ $\mathcal{A}'$ with its associated transition variables in symbolic
  form.}\label{fig:aut_a_annotated}
    \Description[A 3-state automaton with equations on its edges]{
      A 3-state automaton with a self-loop on any character on the
      initial state labelled 1, a transition on c to the accepting state
      and one on d to a state named D, labelled 0 and 1 respectively,
      then one transition from D to FA, the accepting state on d labelled 1.
      The accepting state has a self-loo labelled $l_A -3$.}
  \vspace*{1.9ex}
\end{subfigure}\hfill%
\begin{subfigure}{0.58\textwidth}
  \centering
  \includegraphics[width=\textwidth]{b_annotated}
  \caption{$\mathcal{B}$ with its associated transition variables in symbolic forms.
  The large number of implicitly existentially quantified
  variables~$x_i$ on
  transitions suggest that this automaton has a more complex structure
  with relation to its (free) target variable representing the string length
  than $\mathcal{A}'$. %The variables used here are fresh, and do not correspond to the
  % ones previously used.
}\label{fig:aut_b_annotated}
    \Description[A 4-state automaton with its transitions annotated
    with equations]{A 4-state automaton with transitions from the initial state to B on b,
    one to A on a, labelled $x_{12}$ and $x_{11}$ respectively. State B has one self-loop
    on c labelled $x_9$, one transition back to the initial state labelled 
    b / $x_{11} + x_{12} - 1$. Finally, B has a transition to the accepting state on c,
    labelled $1 0 x_{11}$. A has one self-loop on c labelled $l_A - x_9 - 2x_{11} - 2x_{12}$,
    and one transition on a to the accepting state labelled $x_{11}$.}
\end{subfigure}
\caption{The automata after arithmetic simplification.}\label{fig:propagated}
\end{figure}

### 2024-03-07 extensions.tex
Make automatas bigger, fix incorrect claim

  \begin{algorithm}
    \DontPrintSemicolon
    \caption{$\FindImage{}(\Automaton_1 \times \ldots \times \Automaton_k, \Map)$ will find the Presburger form for the product $\Automaton_1 \times \ldots \times \Automaton_k$ modulo a homomorphism $\Map$ where the only free variable is/are the one(s) representing the monoid element of $\Map$.}\label{alg:find-image}
    \KwData{$\Automaton$, a product from \Calculus{}, a model $a$ assigning counts to the terms that $\Filter$ associate with each transition of $\Automaton$, and our homomorphism $\Map$ that we want to compute the image modulo.}
    \KwResult{a quantifier-free Presburger formula $P$ representing a partial $\Map$-homomorphic image}
    \SetKwFunction{NewTheoremProver}{newTheoremProver}
    \SetKwFunction{EliminateQuantifiers}{eliminateQuantifiers}
    \SetKwFunction{FreshVariable}{freshVariable}
    \SetKwFunction{Assert}{assert}
    \SetKwFunction{GetModel}{getModel}
    \SetKwData{ImageVar}{image}

>>>>

  \begin{algorithm}
    \DontPrintSemicolon{}
    \caption{$\FindImage{}(\Automaton_1 \times \cdots \times \Automaton_k, \Map)$ will find the Presburger form for the product $\Automaton_1 \times \ldots \times \Automaton_k$ modulo a homomorphism $\Map$ where the only free variable is/are the one(s) representing the monoid element of $\Map$.}\label{alg:find-image}
    \KwData{$\Automaton$, a product from \Calculus{}, a model $a$ assigning counts to the terms that $\Filter$ associate with each transition of $\Automaton$, and our homomorphism $\Map$ that we want to compute the image modulo.}
    \KwResult{a quantifier-free Presburger formula $P$ representing a partial $\Map$-homomorphic image}
    \SetKwFunction{NewTheoremProver}{newTheoremProver}
    \SetKwFunction{EliminateQuantifiers}{eliminateQuantifiers}
    \SetKwFunction{FreshVariable}{freshVariable}
    \SetKwFunction{Assert}{assert}
    \SetKwFunction{GetModel}{getModel}
    \SetKwData{ImageVar}{image}
